Traffic Collisions in Toronto: Analysis of Key Variables

### 2024-09-27 paper/paper.qmd
update paper and citations

---
title: "Traffic Collisions in Toronto: Analysis of Key Variables"
subtitle: "An Analysis of Traffic Collision Data from 2014-2024"
author: 
  - Andy Jiang
thanks: "Code and data are available at: https://github.com/AndyYanxunJiang/toronto-traffic-collisions.git"
date: 26 September 2024
date-format: long
abstract: "This paper analyzes traffic collisions in Toronto from 2014 to 2024, focusing on patterns across months, days of the week, times of day, and types of collisions, including injuries, property damage, and failed to remain incidents. Fatalities are analyzed separately as a distinct trend. The data, sourced from Open Data Toronto, reveals a light W-shaped trend in monthly collisions, with peaks in January and June. Collisions are most frequent between 3 PM and 5 PM, with Friday being the highest day for collisions and Sunday the lowest. Property damage accounts for about 71% of all incidents. The number of fatalities has generally declined over the years, though trends fluctuate annually. These findings highlight the need for ongoing road safety initiatives and targeted interventions during high-risk periods to improve long-term traffic safety."
format: pdf
bibliography: references.bib
---

@@

library(tidyverse)
library(knitr)
library(opendatatoronto)
library(dplyr)

@@

This paper examines traffic collisions in Toronto between 2014 and 2024. It aims to identify trends in monthly, daily, and hourly collision occurrences and assess the proportion of different collision types, including injuries, property damage, and failed to remain incidents. Fatalities are analyzed separately as a distinct trend. Analyzing these trends can guide the development of more effective road safety policies and public awareness campaigns.

@@

# Data {#sec-data}
The data used in this paper came from the OpenData Toronto portal through the library `opendatatoronto` [@citeODT], specifically the Police Annual Statistical Report - Traffic Collisions [@CollisionDataCite]. Data were cleaned and analyzed using the open source statistical programming language R [@citeR]. Libraries `tidyverse` [@CiteTD], `knitr` [@knitr] and `dplyr` [@citedplyr] were used for simulating, cleaning and testing. Graphics were made using `ggplot2` [@citeGG].

@@

@tbl-1 provides a random sample of the cleaned dataset, showing key variables such as the occurrence time and type of collision:

```{r}
#| label: tbl-1
#| tbl-cap: Sample of the cleaned collision data
#| echo: false

@@

## Collision Types
@fig-collisions-type shows the proportion of each collision type. Property damage collisions dominate, accounting for approximately 71% of all recorded incidents. This suggests that most collisions involve vehicle-to-vehicle impacts or damage to roadside objects, without causing significant injuries. Failed to remain and injury collisions are relatively similar in proportion, each making up a smaller share of total collisions. Although these types can overlap (e.g., a failed-to-remain collision may also involve injury), the overlap is small enough to justify using a pie chart for this analysis. A more detailed analysis, using a Venn diagram, could account for these overlaps, but the overall trend remains clear: property damage is the most frequent outcome of traffic collisions.

@@

A smaller proportion of collisions involve injuries, where physical harm occurs to drivers or passengers. While these are slightly fewer in number compared to failed-to-remain collisions, they still represent a more severe subset of accidents that require further attention. Failed-to-remain collisions, where drivers flee the scene of the accident, are more common than injury collisions but have the least overlap with both injury and property damage incidents. This minimal overlap suggests that hit-and-run cases are a distinct category of accidents, often disconnected from the other two types, which might require different preventive strategies—such as better surveillance systems or harsher penalties for offenders.

>>>>

---
title: "Traffic Collisions in Toronto: Analysis of Key Variables"
subtitle: "An Analysis of Traffic Collision Data from 2014-2024"
author: 
  - Andy Jiang
thanks: "Code and data are available at: https://github.com/AndyYanxunJiang/toronto-traffic-collisions"
date: today
date-format: long
abstract: "This paper analyzes traffic collisions in Toronto from 2014 to 2024, focusing on patterns across months, days of the week, times of day, and types of collisions, including injuries, property damage, and failed to remain incidents. Fatalities are analyzed separately as a distinct trend. The data, sourced from Open Data Toronto, reveals a light W-shaped trend in monthly collisions, with peaks in January and June. Collisions are most frequent between 3 PM and 5 PM, with Friday being the highest day for collisions and Sunday the lowest. Property damage accounts for about 71% of all incidents. The number of fatalities has generally declined over the years, though trends fluctuate annually. These findings highlight the need for ongoing road safety initiatives and targeted interventions during high-risk periods to improve long-term traffic safety."
format: pdf
bibliography: references.bib
---

@@

library(tidyverse)
library(knitr)
library(opendatatoronto)
library(dplyr)
library(here)

@@

This paper examines traffic collisions in Toronto between 2014 and 2024, a decade that has seen substantial changes in road usage, vehicle technology, and city infrastructure. Despite the growing availability of traffic data, there is still a need for detailed analysis of collision patterns over time and by type. This paper seeks to fill this gap by identifying trends in monthly, daily, and hourly collision occurrences, and assessing the proportion of different collision types, including injuries, property damage, and failed to remain incidents. Fatalities are analyzed separately as a distinct trend. These insights can inform more effective road safety policies, particularly during high-risk periods.

@@

# Data {#sec-data}
The data used in this paper came from the Open Data Toronto portal through the library `opendatatoronto` [@citeODT], specifically the Police Annual Statistical Report - Traffic Collisions [@CollisionDataCite]. Data were cleaned and analyzed using the open source statistical programming language R [@citeR]. Libraries `tidyverse` [@citeTD], `knitr` [@citeknitr], `tibble` [@citetibble], `readr` [@citereadr], and `dplyr` [@citedplyr] were used for simulating, cleaning and testing. The package `here` [@citehere] was utilized for managing file paths to ensure reproducibility. Graphics were made using `ggplot2` [@citeGG].

## Measurement
The traffic collision data is collected by the Toronto Police and made publicly available through Open Data Toronto, where it is graded as Gold with a 100% data quality score. Each entry reflects a documented collision incident, capturing details like time, location, collision type, and vehicle type. However, the dataset does not include information on weather conditions or road quality, which could provide additional context for analyzing collision trends.

@@

@tbl-1 provides a random sample of the cleaned dataset, displaying all relevant variables used for analysis:

```{r}
#| label: tbl-1
#| tbl-cap: Random sample of the cleaned collision data
#| echo: false

@@

## Collision Types
@fig-collisions-type shows the proportion of each collision type. Property damage collisions dominate, accounting for approximately 71% of all recorded incidents. This suggests that most collisions involve vehicle-to-vehicle impacts or damage to roadside objects, without causing significant injuries. Failed to remain and injury collisions are relatively similar in proportion, each making up a smaller share of total collisions. Although these types can overlap (e.g., a failed to remain collision may also involve injury), the overlap is small enough to justify using a pie chart for this analysis. A more detailed analysis, using a Venn diagram, could account for these overlaps, but the overall trend remains clear: property damage is the most frequent outcome of traffic collisions.

@@

A smaller proportion of collisions involve injuries, where physical harm occurs to drivers or passengers. While these are slightly fewer in number compared to failed to remain collisions, they still represent a more severe subset of accidents that require further attention. Failed to remain collisions, where drivers flee the scene of the accident, are more common than injury collisions but have the least overlap with both injury and property damage incidents. This minimal overlap suggests that hit-and-run cases are a distinct category of accidents, often disconnected from the other two types, which might require different preventive strategies—such as better surveillance systems or harsher penalties for offenders.
